The perceptual dimensions of \\ sonority-driven epenthesis

### 2013-09-06 paper.tex
comment
Added Chaha coda cluster section including comparison with other approaches

\title{The perceptual dimensions of \\ sonority-driven epenthesis}
\author{Michelle A. Fullwood}
\date{July 2013}
\begin{document}

@@

\subsection{Another dimension of sonority contour faithfulness}

@@

\subsection{Data: CC clusters}

@@

  The variability therefore lies in (1) whether epenthesis between obstruents is possible (between stops, between a stop and a fricative, and between two fricatives), and (2) whether [r]-sonorant can be broken up by epenthesis or not.
  The variable setting of these two options corresponds to four attested idiolects:

  \ex. \label{idiolects} The four idiolects:
       \a. \label{idiolecta} \tickYes OVO, \tickNo rVN \newline % \textsc{SonSeq} $\gg$ \textsc{SyllCon} $\gg$ \textsc{Dep-IO} $\gg$ \textsc{*R-Son} \newline
           [\textipa{k1rm}], [\textipa{s1B1r}], [\textipa{k1ft}], [\textipa{k1t1f}], [\textipa{s1g1d}]
       \b. \label{idiolectb} \tickNo OVO, \tickNo rVN \newline %\label{idiolectb} \textsc{SyllCon} $\gg$ \textsc{Dep-IO} $\gg$ \textsc{SonSeq, *R-Son} \newline
           [\textipa{k1rm}], [\textipa{s1B1r}], [\textipa{k1ft}], [\textipa{k1tf}], [\textipa{s1gd}]
       \c. \label{idiolectc} \tickYes OVO, \tickYes rVN \newline %\label{idiolectc} \textsc{SonSeq, *R-Son} $\gg$ \textsc{SyllCon} $\gg$ \textsc{Dep-IO} \newline
           [\textipa{k1r1m}], [\textipa{s1B1r}], [\textipa{k1ft}], [\textipa{k1t1f}], [\textipa{s1g1d}]
       \d. \label{idiolectd} \tickNo OVO, \tickYes rVN \newline %\label{idiolectd} \textsc{*R-Son} $\gg$ \textsc{SyllCon} $\gg$ \textsc{Dep-IO} $\gg$ \textsc{SonSeq} \newline
           [\textipa{k1r1m}], [\textipa{s1B1r}], [\textipa{k1ft}], [\textipa{k1tf}], [\textipa{s1gd}]

The evidence of the epenthetic behaviour of coda clusters accords with the {\sc Sonority Angle} prediction that RN should be the most likely to epenthesise out of the falling sonority clusters.

\subsection{Data: CCC clusters}

@@

As with coda clusters, there is considerable variability in epenthetic positioning between idiolects.
In what follows, I assume that the primary forms presented in Section 4.3 of \citet{rose.2000} are preferred forms, and the secondary variants presented in Section 5 are less preferred.

@@

>>>>

\title{The perceptual dimensions of \\ sonority-driven epenthesis}
\author{Michelle A. Fullwood}
\date{Generals Paper, MIT \\ September 2013}
\begin{document}

@@

\subsection{Another dimension of sonority contour faithfulness} \label{depson}

@@

\subsection{CC clusters}

\subsubsection{Data}

@@

The variability therefore lies in (1) whether epenthesis between obstruents is possible (between stops, between a stop and a fricative, and between two fricatives), and (2) whether [r]-sonorant can be broken up by epenthesis or not.

The variable setting of these two options corresponds to four attested idiolects:

  \ex. \label{idiolects} The four idiolects: % TODO: ref
       \a. \label{idiolecta} \tickYes OVO, \tickNo rVN \newline % \textsc{SonSeq} $\gg$ \textsc{SyllCon} $\gg$ \textsc{Dep-IO} $\gg$ \textsc{*R-Son} \newline
           [\textipa{k1rm}], [\textipa{s1B1r}], [\textipa{k1ft}], [\textipa{k1t1f}], [\textipa{s1g1d}]
       \b. \label{idiolectb} \tickYes OVO, \tickYes rVN \newline %\label{idiolectc} \textsc{SonSeq, *R-Son} $\gg$ \textsc{SyllCon} $\gg$ \textsc{Dep-IO} \newline
           [\textipa{k1r1m}], [\textipa{s1B1r}], [\textipa{k1ft}], [\textipa{k1t1f}], [\textipa{s1g1d}]
       \c. \label{idiolectc} \tickNo OVO, \tickNo rVN \newline %\label{idiolectb} \textsc{SyllCon} $\gg$ \textsc{Dep-IO} $\gg$ \textsc{SonSeq, *R-Son} \newline
           [\textipa{k1rm}], [\textipa{s1B1r}], [\textipa{k1ft}], [\textipa{k1tf}], [\textipa{s1gd}]
       \d. \label{idiolectd} \tickNo OVO, \tickYes rVN \newline %\label{idiolectd} \textsc{*R-Son} $\gg$ \textsc{SyllCon} $\gg$ \textsc{Dep-IO} $\gg$ \textsc{SonSeq} \newline
           [\textipa{k1r1m}], [\textipa{s1B1r}], [\textipa{k1ft}], [\textipa{k1tf}], [\textipa{s1gd}]

\subsubsection{Sonority Angle Analysis}

Under a sonority contour faithfulness analysis, epenthesis is triggered by a markedness constraint on the structural conditions of the cluster, not by the markedness of the underlying sonority contour profile itself. Undominated {\sc *ComplexOnset} accounts for the non-existence of initial clusters. {\sc *Coda} is ranked low, and therefore medial clusters are left intact.{\sc *ComplexCoda}, dominated by an {\sc Ident(SonAngle)} constraint, explains why some coda clusters are broken up by epenthesis while others are not.

In idiolect \ref{idiolecta}, all coda clusters with a {\sc Sonority Angle} smaller than 1.5 -- the rising and level sonority clusters -- are broken up by epenthesis, while those with a {\sc Sonority Angle} greater than 1.5 -- the falling sonority clusters -- are left intact.

%TODO: diagram

The constraint ranking for this idiolect is thus:

\ex. {\sc Ident(SonAngle)}$<$1.5 $\gg$ {\sc *ComplexCoda}

\begin{center} \renewcommand*\arraystretch{1.2}
\scalebox{1}[1]{\begin{tabular}[t]{|rrl||c|c|} \hline 
\multicolumn{3}{|c||}{Input:~/\textipa{k1tf}/} & {\sc Ident(SonAngle)}$<$1.5 & {\sc *ComplexCoda} \\[0.5ex]
\hline \hline a. & & \textipa{k1tf} & & $\ast$! \\
\hline b. & \ding{43} & \textipa{k1t1f} & 0.59 & \\
\hline \end{tabular}} \renewcommand*\arraystretch{1} \end{center}

\begin{center} \renewcommand*\arraystretch{1.2}
\scalebox{1}[1]{\begin{tabular}[t]{|rrl||c|c|} \hline 
\multicolumn{3}{|c||}{Input:~/\textipa{k1rm}/} & {\sc Ident(SonAngle)}$<$1.5 & {\sc *ComplexCoda} \\[0.5ex]
\hline \hline c. &  \ding{43} & \textipa{k1rm} & & $\ast$ \\
\hline d. & & \textipa{k1r1m} & 1.89 $\ast$! & \\
\hline \end{tabular}} \renewcommand*\arraystretch{1} \end{center}

This approach does make one incorrect prediction: it predicts that /\textipa{Bm}/ should not undergo epenthesis under this ranking, since /\textipa{B}N/ has a {\sc Sonority Angle} of 1.65. But it always does:

\ex. /\textipa{s@B-m}/ $\rightarrow$ \textipa{s@B1m} `people-{\sc emph}'  % TODO: ref

The only clusters that give us evidence of this are /\textipa{Bm}/ clusters. These are not seen root-internally due to a high-ranking OCP effect. % TODO: ref: Greenberg cited in Rose
I suggest that there might be an undominated markedness constraint against this cluster -- either OCP(labial) or another constraint based on articulatory or perceptual difficulty -- that outranks all relevant faithfulness constraints.

Proceeding to idiolect \ref{idiolectb}, we find that [r]-sonorant clusters are allowed word-finally. Because {\sc Sonority Angle} permits a linear separation of the [r]-sonorant clusters from the remainder of the falling sonority clusters, this is straightforward to explain
by the movement of the cut-off line from 1.5 to 1.9.

% TODO: diagram

\ex. {\sc Ident(SonAngle)}$<$1.9 $\gg$ {\sc *ComplexCoda}

\begin{center} \renewcommand*\arraystretch{1.2}
\scalebox{1}[1]{\begin{tabular}[t]{|rrl||c|c|} \hline 
\multicolumn{3}{|c||}{Input:~/\textipa{k1rm}/} & {\sc Ident(SonAngle)}$<$1.5 & {\sc *ComplexCoda} \\[0.5ex]
\hline \hline a. &   & \textipa{k1rm} & & $\ast$! \\
\hline b. & \ding{43} & \textipa{k1r1m} & 1.89 & \\
\hline \end{tabular}} \renewcommand*\arraystretch{1} \end{center}

Idiolects \ref{idiolectc} and \ref{idiolectd} are identical to \ref{idiolecta} and \ref{idiolectb} respectively, except that obstruent-obstruent clusters do not undergo epenthesis under any circumstances. This can be achieved by ranking {\sc Dep}(+son)
above {\sc *ComplexCoda}.

\ex. Ranking for idiolect \ref{idiolectc}

\begin{center} \renewcommand*\arraystretch{1.2}
\scalebox{1}[1]{\begin{tabular}[t]{|rrl||c:c|c|} \hline 
\multicolumn{3}{|c||}{Input:~/\textipa{s1gd}/} & {\sc Dep}(+son) & {\sc Ident(SonAngle)}$<$1.5 & {\sc *ComplexCoda} \\[0.5ex]
\hline \hline a. & \ding{43} & \textipa{s1gd} & & & \cellcolor{lightgray}$\ast$ \\
\hline b. & & \textipa{s1g1d} & $\ast$! & 1.37 & \cellcolor{lightgray} \\
\hline \end{tabular}} \renewcommand*\arraystretch{1} \end{center}

To summarise, coda epenthesis behaviour in Chaha can be explained by the ranking of a {\sc Ident(SonAngle)}$<$n constraint above the markedness constraint triggering epenthesis, {\sc *ComplexCoda}. The evidence of the epenthetic behaviour of Chaha coda clusters accords with the {\sc Sonority Angle} prediction that RN should be the most likely to epenthesise out of the falling sonority clusters. It also demonstrates the action of a faithfulness constraint against introducing a sonorant region between two obstruents, {\sc Dep}(+son), as defined in $\S$\ref{depson}.

\ex. Rankings for the four idiolects:
     \a. {\sc Ident(SonAngle)}$<$1.5 $\gg$ {\sc *ComplexCoda} $\gg$ {\sc Dep}(+son)
     \b. {\sc Ident(SonAngle)}$<$1.9 $\gg$ {\sc *ComplexCoda} $\gg$ {\sc Dep}(+son)
     \c. {\sc Ident(SonAngle)}$<$1.5, {\sc Dep}(+son) $\gg$ {\sc *ComplexCoda}
     \c. {\sc Ident(SonAngle)}$<$1.9, {\sc Dep}(+son) $\gg$ {\sc *ComplexCoda}

\subsubsection{Comparison to other approaches}

{\sc Sonority Rise} can explain idiolects \ref{idiolecta} and \ref{idiolectc}, where the cut-off is drawn between rising and level sonority clusters on the one hand and falling sonority clusters on the other, it is less straightforward to explain the behaviour of idiolects \ref{idiolectb} and \ref{idiolectd}. This is because the cluster hierarchy predicted by {\sc Sonority Rise} does not permit a linear separation of [r]-sonorant clusters from the other falling sonority clusters.

% TODO: diagram

\citet{rose.2000} offers an explanation for these facts in terms of markedness of the sonority contours of the surface clusters. Medial clusters are subject to Syllable Contact, which bans rising sonority clusters across a syllable boundary, while coda clusters are subject to Sonority Sequencing.

The fact that medial clusters are never broken up by epenthesis even when they violate Syllable Contact is due to a constraint against the configuration [V.C\textipa{1}.CV], {\sc *MedialLight}. This constraint is unnecessary under my analysis as the only constraint that would militate against medial clusters, {\sc *Coda}, is not active.

In addition, Sonority Sequencing does not allow a divide of the [r]-sonorant clusters versus the remainder of the clusters, whether in Rose's categorical formulation of Sonority Sequencing or a family of {\sc *Dis} constraints.

% TODO: diagram

Therefore, Rose needs to invoke the following constraint to cause epenthesis in these clusters:

\ex. {\sc *R-Sonorant}: no [r]-sonorant sequences \citep[(41)]{rose.2000}

For those speakers that permit obstruent-obstruent clusters in defiance of Sonority Sequencing, Rose ranks Sonority Sequencing low, using the following constraint to trigger epenthesis in other clusters instead:

\ex. {\sc *C-Son\#}: a word-final sonorant consonant must be preceded by a vowel or [r].

Rose's approach thus requires the formulation of two constraints specifically targeting [r]. The advantage of doing so is that it avoids the difficulty of forbidding [\textipa{Bm}] clusters which we encountered above. On the other hand, the sonority contour faithfulness approach can get away without specifically mentioning [r] at all, by virtue of the {\sc Sonority Angle} hierarchy being able to group [r]-sonorant clusters together with the level and rising sonority clusters, at the expense of requiring a constraint against [\textipa{Bm}] clusters.

\subsection{CCC clusters}

\subsubsection{Data and Analysis}

@@

As with coda clusters, there is considerable inter- and intra-speaker variability in epenthetic positioning between idiolects. The following data is from the main idiolect discussed in Section 4.3 of \citep{rose.2000}. Underlined forms have alternate forms in other idiolects.

In

@@

% TODO
